Multivariate Analysis

### 2021-02-18 multivariate.Rmd
cleaned up and added an imputed RF

---
title: "Multivariate Analysis"
author: "Scott Coffin"
date: "1/19/2021"
output: html_document
---
# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(tidyverse) #General everything
library(RColorBrewer)
library(ggplot2) #General plotting
library(ggrepel) #For adding text labels that repel away from data points
library(calecopal) #Color palette
library(shiny) #Runs shiny
library(shinythemes) #Shiny theme for the page
library(shinyWidgets) #Widgets
library(scales) #SSD - Use the percent format
library(reshape2) #Overview tab - melts bars together
library(ssdtools) #SSD package
library(DT) #Build HTML data tables
#library(plotly) #Make plots interactive
library(viridis) #Colors
library(scales) #To use "percent" function
library(shinyjs) #Exploration tab - reset button
library(tigerstats) #turns things into percents
library(ggbeeswarm) #plot all points
library(fitdistrplus) #alt SSD 
library(RColorBrewer) #colors
library(pheatmap) #pretty heat maps
library(rpart)  #for trees
#library(rattle)    # Fancy tree plot This is a difficult library to install (https://gist.github.com/zhiyzuo/a489ffdcc5da87f28f8589a55aa206dd) 
library(rpart.plot)             # Enhanced tree plots
library(RColorBrewer)       # Color selection for fancy tree plot
library(party)                  # Alternative decision tree algorithm
# library(partykit)               # Convert rpart object to BinaryTree
library(pROC)   #for ROC curves
library(uwot) #umap
library(ISLR)  #for the Carseat Data
```
## Data import
```{r data import}
# Load finalized dataset.
aoc <- read_csv("AquaticOrganisms_Clean_final.csv", guess_max = 10000)

@@

  mutate(dose.particles.mL.master.converted.reported = factor(dose.particles.mL.master.converted.reported))

@@

multiVar <- aoc_z %>% dplyr::select(#doi, size.category, 
                                    size_f,
                                    size.length.um.used.for.conversions, 
                                    shape, 
                                    polymer, 
                                    particle.volume.um3, 
                                    density.mg.um.3, 
                                    organism.group,
                                    environment, 
                                    bio.org, #biological level of organization
                                    #af.time, #assessment factor based on exposure time
                                    treatments, #number of doses (no including control)
                                    effect, #yes no
                                    exposure.duration.d, 
                                    exposure.route, #Factor
                                    organism.group, #factor
                                    media.temp, #numeric
                                    lvl1_f, #endpoints
                                    lvl2_f, #endpoints
                                   # lvl3, 
                                     dose.mg.L.master, 
                                    sex, #factor
                                    media.ph, #numeric
                                    media.sal.ppt, #numeric
                                    dose.particles.mL.master,
                                   effect.metric, #NOEC LOEC
                                    functional.group, #factor
                                    charge, #positive or negatibe
                                    zetapotential.mV, # numeric   
                                   max.size.ingest.mm) %>%  #max ingestible size
  filter(!size_f == "Not Reported")  #take out not reported

#recode variables
multiVar <- multiVar %>% mutate(effect_10 = case_when(
    effect == "Y" ~ 1,
    effect == "N" ~ 0
  ))# %>% 
  #mutate_all(is.character, ~as.factor())

@@

# Analysis
## Data Exploration
### Completeness vby Size
```{r completeness heatmap}
CompletenessSummary_size <- multiVar %>%
  group_by(size_f) %>% 
     summarise_all((name = ~sum(is.na(.))/length(.))) %>% 
  mutate(across(is.numeric,~round(., 2))) %>% 
  mutate(across(is.numeric, ~100 *(1 -.)))
CompletenessSummary_size
```

```{r}
require()
#convert to matrix and transpose
transposed_size <- as.data.frame(t(as.matrix(CompletenessSummary_size[2:25]))) %>% #1 is category, 2-6 are variables
  arrange('1nm < 100nm')
#reassign column names
colnames(transposed_size) <- c("1nm < 100nm", "100nm < 1µm", "1µm < 100µm", "100µm < 1mm", "1mm < 5mm")
#transposedTag
#format as matrix
MissingMatrix_size <- data.matrix(transposed_size)
#build heatmap
pheatmap(MissingMatrix_size,
                           main = "Data Completeness by Size Bin", #title
                           fontsize = 12,
                           cluster_rows = FALSE, cluster_cols = FALSE,#disable dendrograms
                           display_numbers = TRUE,
                           treeheight_row = 0, treeheight_col = 0, #keeps clustering after dropping dendrograms
                           col = rev(brewer.pal(n = 9, name = "PuBu"))) #blue color scheme with 9 colors)

```
### Completeness by Environment
```{r completeness heatmap}
CompletenessSummary_environment <- multiVar %>%
  filter(!environment == "NA") %>% 
  group_by(environment) %>% 
     summarise_all((name = ~sum(is.na(.))/length(.))) %>% 
  mutate(across(is.numeric,~round(., 2))) %>% 
  mutate(across(is.numeric, ~100 *(1 -.)))
CompletenessSummary_environment
```

```{r}
require()
#convert to matrix and transpose
transposed_environment <- as.data.frame(t(as.matrix(CompletenessSummary_environment[2:25])))
#reassign column names
colnames(transposed_environment) <- c("Freshwater", "Marine", "Terrestrial")
#transposedTag
#format as matrix
MissingMatrix_environment <- data.matrix(transposed_environment)
#build heatmap
pheatmap(MissingMatrix_environment,
                           main = "Data Completeness by environment", #title
                           fontsize = 12,
                           cluster_rows = FALSE, cluster_cols = FALSE,#disable dendrograms
                           display_numbers = TRUE,
                           treeheight_row = 0, treeheight_col = 0, #keeps clustering after dropping dendrograms
                           col = rev(brewer.pal(n = 9, name = "PuBu"))) #blue color scheme with 9 colors)

@@

```{r}
## Estimate an OLS Regression
fitols <- lm(effect_10 ~ size.length.um.used.for.conversions + particle.volume.um3 + exposure.duration.d + media.temp + dose.particles.mL.master +mouth, 
             na.action = na.omit, 
             data = multiVar)
summary(fitols)
```

@@

## PCA

@@

```{r}
# prune the tree
pt1 <- prune(t1,cp)
#pt1<- prune(t1, cp= t1$cptable[which.min(t1$cptable[,"xerror"]),"CP"])

# plot the pruned tree
plot(pt1, uniform=TRUE,
   main="Pruned Classification Tree");text(pt1, use.n=TRUE, all=TRUE, cex=.8)

#post(pfit, file = "c:/ptree.ps",
 #  title = "Pruned Classification Tree for Kyphosis")
```
```{r}
#find proportion of correct predictions using test set
t1_pruned_predict <- predict(pt1, testSet, type="class")
mean(t1_pruned_predict == testSet$effect)
#
```
This is not an improvement over an unprunend tree.. We need to check that this holds up for different training and test sets. This is easily done by creating multiple random partitions of the dataset and checking the efficacy of pruning for each. To do this efficiently, I’ll create a function that takes the training fraction, number of runs (partitions) and the name of the dataset as inputs and outputs the proportion of correct predictions for each run. It also optionally prunes the tree.

```{r}
#function to do multiple runs
multiple_runs_classification <- function(train_fraction,n,dataset,prune_tree=FALSE){
fraction_correct <- rep(NA,n)
set.seed(42)
for (i in 1:n){
  dataset[,"train"] <- ifelse(runif(nrow(dataset))<0.8,1,0)
  trainColNum <- grep("train",names(dataset))
  typeColNum <- grep("effect",names(dataset))
  trainSet <- dataset[dataset$train==1,-trainColNum]
  testSet <- dataset[dataset$train==0,-trainColNum]
  rpart_model <- rpart(effect~ size.length.um.used.for.conversions + particle.volume.um3 + exposure.duration.d + media.temp + dose.particles.mL.master,data = trainSet, method="class")
if(prune_tree==FALSE) {
  rpart_test_predict <- predict(rpart_model,testSet[,-typeColNum],type="class")
  fraction_correct[i] <- mean(rpart_test_predict==testSet$effect)
  }else{
    opt <- which.min(rpart_model$cptable[,"xerror"])
    cp <- rpart_model$cptable[opt, "CP"]
    pruned_model <- prune(rpart_model,cp)
    rpart_pruned_predict <- predict(pruned_model,testSet[,-typeColNum],type="class")
    fraction_correct[i] <- mean(rpart_pruned_predict == testSet$effect)
  }
  }
return(fraction_correct)
}
```

Note that in the above,  I have set the default value of the prune_tree to FALSE, so the function will execute the first branch of the if statement unless the default is overridden.

OK, so let’s do 50 runs with and without pruning, and check the mean and variance of the results for both sets of runs.
```{r}
#50 runs, no pruning
unpruned_set <- multiple_runs_classification(0.8,50,multiVar_sub)
mean(unpruned_set)
#[1] 0.7261882
sd(unpruned_set)
#[1] 0.01554347
#50 runs, with pruning
pruned_set <- multiple_runs_classification(0.8,50,multiVar_sub,prune_tree=TRUE)
mean(pruned_set)
#[1] 0.7261875
sd(pruned_set)
#[1] 0.01552285
```

@@

```{r fit results}
train2 <- multiVar_sub %>% dplyr::select(size.length.um.used.for.conversions,
                                  particle.volume.um3,exposure.duration.d,media.temp,dose.particles.mL.master)
train3 <-multiVar_sub %>% dplyr::select(size.length.um.used.for.conversions,
                                  particle.volume.um3,exposure.duration.d,media.temp,dose.particles.mL.master,
                                  effect_10)

@@

```{r}
#Obtaining predicted probabilites for Test data
tree.probs=predict(crf,
                 newdata = train2,
                 type="prob")

#Calculate ROC curve
rocCurve.tree <- roc(train3$effect_10,tree.probs[,'1'])
#plot the ROC curve
plot(rocCurve.tree,col=c(4))
```

Alternative ROC Curve
```{r}
require(pROC)
predicted <- predict(crf, train2, OOB=TRUE, type= "response")
auc(as.numeric(train3$effect_10), as.numeric(predicted))
#Calculate ROC curve
rocCurve.tree <- roc(train3$effect_10,as.numeric(predicted))
#plot the ROC curve
plot(rocCurve.tree,col=c(4))
```
```{r}
#plot reciever operating curve
perf <- performance(preds,"tpr","fpr")
```

@@

```{r}
crf.prob <- matrix(unlist(crfsrvy.prob), ncol=2, byrow=TRUE)
apply(crf.prob,2,mean)
#[1] 0.2524514 0.7475486
tab <- round(cbind(by(rpart.prob[,2], INDICES=t1$where, mean), by(crf.prob[,2], INDICES=t1$where, mean)), 4)
colnames(tab) <- c("rpart", "cforest")
kable(tab)
```
# Non-linear Dose Response
If we hold all paramaters constant except for dose, we should observe a non-linear dose-response behavior that can be best predicted by a 3- or 4-parameter Hill Equation. The most available information is likely for polystyrene spheres in crustaceans.There are 21 studies available.

>>>>

---
title: "Multivariate Analysis"
author: "Scott Coffin"
date: "1/19/2021"
output: 
  html_document:
    code_folding: hide
    theme: sandstone
    toc: yes
    toc_float: yes
    toc_depth: 4
    number_sections: true
  word_document:
    toc: yes
---
```{r global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      warning=FALSE, message=FALSE, echo = FALSE)
```
---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(tidyverse) #General everything
library(RColorBrewer)
library(ggplot2) #General plotting
library(ggrepel) #For adding text labels that repel away from data points
library(calecopal) #Color palette
library(shiny) #Runs shiny
library(shinythemes) #Shiny theme for the page
library(shinyWidgets) #Widgets
library(scales) #SSD - Use the percent format
library(reshape2) #Overview tab - melts bars together
library(ssdtools) #SSD package
library(DT) #Build HTML data tables
#library(plotly) #Make plots interactive
library(viridis) #Colors
library(scales) #To use "percent" function
library(tigerstats) #turns things into percents
library(ggbeeswarm) #plot all points
library(fitdistrplus, exclude = 'select') #alt SSD 
library(RColorBrewer) #colors
library(pheatmap) #pretty heat maps
library(rpart)  #for trees
#library(rattle)    # Fancy tree plot This is a difficult library to install (https://gist.github.com/zhiyzuo/a489ffdcc5da87f28f8589a55aa206dd) 
library(rpart.plot)             # Enhanced tree plots
library(RColorBrewer)       # Color selection for fancy tree plot
library(party)                  # Alternative decision tree algorithm
# library(partykit)               # Convert rpart object to BinaryTree
library(pROC)   #for ROC curves
library(uwot) #umap
library(ISLR)  #for the Carseat Data
library(lme4) #general linearized mixed model GLMM
library(quantregForest)
library(caret)
library(tidyverse)
library(tidymodels)
library(skimr)
library(sf)
library(ggspatial)
library(nhdplusTools)
library(patchwork)
library(Metrics)
library(gt)
library(randomForest)
```
## Data import
```{r data import}
# Load finalized dataset.
aoc <- read_csv("AquaticOrganisms_Clean_final.csv", guess_max = 10000) %>% 
  mutate(ID = paste0("ID",row_number()))

@@

  mutate(dose.particles.mL.master.converted.reported = factor(dose.particles.mL.master.converted.reported)) %>% 
   mutate(dose.um3.mL.master = particle.volume.um3 * dose.particles.mL.master) %>%   #calculate volume/mL
  mutate(af.time_noNA = replace_na(af.time, "Unavailable")) %>% 
  mutate(acute.chronic_f = factor(case_when(af.time_noNA == 10 ~ "Acute",
                                            af.time_noNA == 1 ~ "Chronic",
                                            af.time_noNA == "Unavailable" ~ "Unavailable"))) %>%    #factorize assesment factor time into chronic/acute
  mutate(dose.mg.L.master.AF.noec = dose.mg.L.master * af.noec) %>% 
  mutate(dose.particles.mL.master.AF.noec = dose.particles.mL.master * af.noec) %>% 
  mutate(effect_f = factor(effect)) %>% 
  mutate_if(is.character, as.factor) %>% 
  mutate(effect_10 = case_when(
     effect_f == "Y" ~ 1,
     effect_f == "N" ~ 0))

@@

multiVar <- aoc_z %>% dplyr::select(#doi, size.category, 
                                    size_f,
                                    size.length.um.used.for.conversions, 
                                    shape, 
                                    polymer, 
                                    particle.volume.um3, 
                                    density.mg.um.3, 
                                    organism.group,
                                    environment, 
                                    bio.org, #biological level of organization
                                    #af.time, #assessment factor based on exposure time
                                    treatments, #number of doses (no including control)
                                    effect, #yes no
                                    effect_f,
                                    effect_10,
                                    size_f,
                                    exposure.duration.d, 
                                    exposure.route, #Factor
                                    organism.group, #factor
                                    media.temp, #numeric
                                    lvl1_f, #endpoints
                                    lvl2_f, #endpoints
                                   # lvl3, 
                                     dose.mg.L.master, 
                                    sex, #factor
                                    media.ph, #numeric
                                    media.sal.ppt, #numeric
                                    dose.particles.mL.master,
                                   effect.metric, #NOEC LOEC
                                    functional.group, #factor
                                    charge, #positive or negatibe
                                    zetapotential.mV, # numeric   
                                   max.size.ingest.mm,#max ingestible size
                                   acute.chronic_f,
                                   dose.mg.L.master.AF.noec,
                                   dose.particles.mL.master.AF.noec,
                                   max.size.ingest.mm, #maximum ingestible size range
                                   effect.score) %>%  #1 = minor, 2 = photosynthesis, feeding, 3 = growth, chlorophyll content, 4 = reproduction, 5  = population growth, 6 = survival
  filter(!size_f == "Not Reported") %>%   #take out not reported 
                                 #  max.size.ingest.mm) %>%  #max ingestible size
  filter(!size_f == "Not Reported")  #take out not reported

#recode variables
# multiVar <- multiVar %>% mutate(effect_10 = case_when(
#     effect == "Y" ~ 1,
#     effect == "N" ~ 0
#   ))# %>% 
#   #mutate_all(is.character, ~as.factor())

@@

# Analysis
## Data Exploration
### Completeness vby Size
```{r completeness heatmap 1}
CompletenessSummary_size <- multiVar %>%
  group_by(size_f) %>% 
     summarise_all((name = ~sum(is.na(.))/length(.))) %>% 
  mutate(across(is.numeric,~round(., 2))) %>% 
  mutate(across(is.numeric, ~100 *(1 -.)))
CompletenessSummary_size
```

```{r}
require(pheatmap)
#convert to matrix and transpose
transposed_size <- as.data.frame(t(as.matrix(CompletenessSummary_size[2:25]))) %>% #1 is category, 2-6 are variables
  arrange('1nm < 100nm')
#reassign column names
colnames(transposed_size) <- c("1nm < 100nm", "100nm < 1µm", "1µm < 100µm", "100µm < 1mm", "1mm < 5mm")
#transposedTag
#format as matrix
MissingMatrix_size <- data.matrix(transposed_size)
#build heatmap
pheatmap(MissingMatrix_size,
                           main = "Data Completeness by Size Bin", #title
                           fontsize = 12,
                           cluster_rows = FALSE, cluster_cols = FALSE,#disable dendrograms
                           display_numbers = TRUE,
                           treeheight_row = 0, treeheight_col = 0, #keeps clustering after dropping dendrograms
                           col = rev(brewer.pal(n = 9, name = "PuBu"))) #blue color scheme with 9 colors)

```
### Completeness by Environment
```{r completeness heatmap 2}
CompletenessSummary_environment <- multiVar %>%
  filter(!environment == "NA") %>% 
  group_by(environment) %>% 
     summarise_all((name = ~sum(is.na(.))/length(.))) %>% 
  mutate(across(is.numeric,~round(., 2))) %>% 
  mutate(across(is.numeric, ~100 *(1 -.)))
CompletenessSummary_environment
```

```{r}
require(pheatmap)
#convert to matrix and transpose
transposed_environment <- as.data.frame(t(as.matrix(CompletenessSummary_environment[2:25])))
#reassign column names
colnames(transposed_environment) <- c("Freshwater", "Marine", "Terrestrial")
#transposedTag
#format as matrix
MissingMatrix_environment <- data.matrix(transposed_environment)
#build heatmap
pheatmap(MissingMatrix_environment,
                           main = "Data Completeness by environment", #title
                           fontsize = 12,
                           cluster_rows = FALSE, cluster_cols = FALSE,#disable dendrograms
                           display_numbers = TRUE,
                           treeheight_row = 0, treeheight_col = 0, #keeps clustering after dropping dendrograms
                           col = rev(brewer.pal(n = 9, name = "PuBu"))) #blue color scheme with 9 colors)

@@

```{r}
## Estimate an OLS Regression
fitols <- lm(effect_10 ~ size.length.um.used.for.conversions + particle.volume.um3 + exposure.duration.d + media.temp + dose.particles.mL.master, 
             na.action = na.omit, 
             data = multiVar)
summary(fitols)
```

@@

# PCA

@@

```{r}
# # prune the tree
# pt1 <- prune(t1,cp)
# pt1<- prune(t1, cp= t1$cptable[which.min(t1$cptable[,"xerror"]),"CP"])
# 
# # plot the pruned tree
# plot(pt1, uniform=TRUE,
#    main="Pruned Classification Tree");text(pt1, use.n=TRUE, all=TRUE, cex=.8)
# 
# #post(pfit, file = "c:/ptree.ps",
#  #  title = "Pruned Classification Tree for Kyphosis")
```
```{r}
# #find proportion of correct predictions using test set
# t1_pruned_predict <- predict(pt1, testSet, type="class")
# mean(t1_pruned_predict == testSet$effect)
# #
```
This is not an improvement over an unprunend tree.. We need to check that this holds up for different training and test sets. This is easily done by creating multiple random partitions of the dataset and checking the efficacy of pruning for each. To do this efficiently, I’ll create a function that takes the training fraction, number of runs (partitions) and the name of the dataset as inputs and outputs the proportion of correct predictions for each run. It also optionally prunes the tree.

```{r}
# <!-- # #function to do multiple runs -->
# <!-- # multiple_runs_classification <- function(train_fraction,n,dataset,prune_tree=FALSE){ -->
# <!-- # fraction_correct <- rep(NA,n) -->
# <!-- # set.seed(42) -->
# <!-- # for (i in 1:n){ -->
# <!-- #   dataset[,"train"] <- ifelse(runif(nrow(dataset))<0.8,1,0) -->
# <!-- #   trainColNum <- grep("train",names(dataset)) -->
# <!-- #   typeColNum <- grep("effect",names(dataset)) -->
# <!-- #   trainSet <- dataset[dataset$train==1,-trainColNum] -->
# <!-- #   testSet <- dataset[dataset$train==0,-trainColNum] -->
# <!-- #   rpart_model <- rpart(effect~ size.length.um.used.for.conversions + particle.volume.um3 + exposure.duration.d + media.temp + dose.particles.mL.master,data = trainSet, method="class") -->
# <!-- # if(prune_tree==FALSE) { -->
# <!-- #   rpart_test_predict <- predict(rpart_model,testSet[,-typeColNum],type="class") -->
# <!-- #   fraction_correct[i] <- mean(rpart_test_predict==testSet$effect) -->
# <!-- #   }else{ -->
# <!-- #     opt <- which.min(rpart_model$cptable[,"xerror"]) -->
# <!-- #     cp <- rpart_model$cptable[opt, "CP"] -->
# <!-- #     pruned_model <- prune(rpart_model,cp) -->
# <!-- #     rpart_pruned_predict <- predict(pruned_model,testSet[,-typeColNum],type="class") -->
# <!-- #     fraction_correct[i] <- mean(rpart_pruned_predict == testSet$effect) -->
# <!-- #   } -->
# <!-- #   } -->
# <!-- # return(fraction_correct) -->
# <!-- # } -->
# <!-- # ``` -->
#  
# <!-- # Note that in the above,  I have set the default value of the prune_tree to FALSE, so the function will execute the first branch of the if statement unless the default is overridden. -->
# <!-- #  -->
# <!-- # OK, so let’s do 50 runs with and without pruning, and check the mean and variance of the results for both sets of runs. -->
# <!-- # ```{r} -->
# <!-- # #50 runs, no pruning -->
# <!-- # unpruned_set <- multiple_runs_classification(0.8,50,multiVar_sub) -->
# <!-- # mean(unpruned_set) -->
# <!-- # #[1] 0.7261882 -->
# <!-- # sd(unpruned_set) -->
# <!-- # #[1] 0.01554347 -->
# <!-- # #50 runs, with pruning -->
# <!-- # pruned_set <- multiple_runs_classification(0.8,50,multiVar_sub,prune_tree=TRUE) -->
# <!-- # mean(pruned_set) -->
# <!-- # #[1] 0.7261875 -->
# <!-- # sd(pruned_set) -->
# <!-- # #[1] 0.01552285 -->
# <!-- # ``` -->
```

@@

```{r fit results 1}
train2 <- multiVar_sub %>% dplyr::select(size.length.um.used.for.conversions,
                                  particle.volume.um3,exposure.duration.d,media.temp,dose.particles.mL.master)
train3 <-multiVar_sub %>% dplyr::select(size.length.um.used.for.conversions,
                                  particle.volume.um3,exposure.duration.d,media.temp,dose.particles.mL.master,
                                  effect_10)

@@

<!-- # ```{r} -->
<!-- # #Obtaining predicted probabilites for Test data -->
<!-- # tree.probs=predict(crf, -->
<!-- #                  newdata = train2, -->
<!-- #                  type="prob") -->
<!-- #  -->
<!-- # #Calculate ROC curve -->
<!-- # rocCurve.tree <- roc(train3$effect_10,tree.probs[,'1']) -->
<!-- # #plot the ROC curve -->
<!-- # plot(rocCurve.tree,col=c(4)) -->
<!-- # ``` -->

Alternative ROC Curve
```{r}
require(pROC)
predicted <- predict(crf, train2, OOB=TRUE, type= "response")
auc(as.numeric(train3$effect_10), as.numeric(predicted))
#Calculate ROC curve
rocCurve.tree <- roc(train3$effect_10,as.numeric(predicted))
#plot the ROC curve
plot(rocCurve.tree,col=c(4))
```

@@

<!-- # ```{r} -->
<!-- # crf.prob <- matrix(unlist(crfsrvy.prob), ncol=2, byrow=TRUE) -->
<!-- # apply(crf.prob,2,mean) -->
<!-- # #[1] 0.2524514 0.7475486 -->
<!-- # tab <- round(cbind(by(rpart.prob[,2], INDICES=t1$where, mean), by(crf.prob[,2], INDICES=t1$where, mean)), 4) -->
<!-- # colnames(tab) <- c("rpart", "cforest") -->
<!-- # kable(tab) -->
<!-- # ``` -->

#GLMM
```{r}
#tutorial https://aosmith.rbind.io/2020/08/20/simulate-binomial-glmm/
require(lme4)

mod = glm(effect_f ~ size.length.um.used.for.conversions + polymer + particle.volume.um3 + density.mg.um.3 + organism.group + bio.org + treatments + effect_f + exposure.duration.d + exposure.route + lvl1_f + dose.mg.L.master, 
            data = aoc_z,
            family = binomial(link = "logit") )
mod
```

```{r}
bin_glmm_fun = function(n_sites = 10,
                        b0 = 0,
                        b1 = 1.735,
                        num_samp = 50,
                        site_var = 0.5) {
     site = rep(LETTERS[1:n_sites], each = 2)
     plot = paste(site, rep(1:2, times = n_sites), sep = "." )
     treatment = rep( c("treatment", "control"), times = n_sites)
     dat = data.frame(site, plot, treatment)

     site_eff = rep( rnorm(n = n_sites, mean = 0, sd = sqrt(site_var) ), each = 2)

     log_odds = with(dat, b0 + b1*(treatment == "treatment") + site_eff)
     prop = plogis(log_odds)
     dat$num_samp = num_samp
     dat$y = rbinom(n = n_sites*2, size = num_samp, prob = prop)

     glmer(cbind(y, num_samp - y) ~ treatment + (1|site),
           data = dat,
           family = binomial(link = "logit") )
}

set.seed(16)
bin_glmm_fun()

# Random Forest Modelling

## Preparation
```{r}
library(quantregForest)
library(caret)
library(tidyverse)
library(tidymodels)
library(skimr)
library(sf)
library(ggspatial)
library(nhdplusTools)
library(patchwork)
library(Metrics)
library(gt)
```

```{r}
aoc_z_2 <- aoc_z %>% 
  mutate_if(is.character, as.factor)# %>%
  # mutate(effect_10 = case_when( #convert ordinal to numeric
  #    effect_f == "Yes" ~ 1,
  #    effect_f == "No" ~ 0))

#choose relevant predictors and log-transform
multiVar_small_log <- aoc_z_2 %>% 
  dplyr::select(size_f,
                ID, #row number ID for split/joining by study
                doi, #need to split studies
                size.length.um.used.for.conversions, shape, polymer, particle.volume.um3, density.mg.um.3, organism.group, bio.org, treatments, effect_f, exposure.duration.d, exposure.route, lvl1_f, dose.mg.L.master) %>% 
  mutate_if(~is.numeric(.) && (.) > 0, log10) %>% 
  drop_na() %>%  #drop missing
 mutate(effect_10 = case_when( #convert ordinal to numeric
     effect_f == "Y" ~ 1,
     effect_f == "N" ~ 0
   ))# %>% 
 #  select(-effect_f)
skim(multiVar_small_log)
```

```{r}
#no log transform for comparison
multiVar_small <- aoc_z %>% 
  dplyr::select(size_f, size.length.um.used.for.conversions, shape, polymer, particle.volume.um3, density.mg.um.3, organism.group, bio.org, treatments, effect_f, life.stage,
                exposure.duration.d, lvl1_f, dose.mg.L.master, dose.particles.mL.master, dose.um3.mL.master)# %>% 
  #drop_na() #%>%  #drop missing
# mutate(effect_10 = case_when( #convert ordinal to numeric
#     effect == "Y" ~ 1,
#     effect == "N" ~ 0
#   )) %>% 
  # select(-effect)

#ensure completeness
skim(multiVar_small)
```

##NON-LOG TRANSFORMED
### Missing Value Imputation by Rough Fix
```{r}
#rough fix
multiVar_small_roughfix <- na.roughfix(multiVar_small)

# Create calibration and validation splits with tidymodels initial_split() function.
set.seed(4)
multiVar_small_split <- multiVar_small_roughfix %>%
  initial_split(prop = 0.75, strata = polymer) # splits data into training and testing set.
# default is 3/4ths split (but 75% training, 25% testing).
# Stratification (strata) = grouping training/testing sets by region, state, etc.
# Using the "strata" call ensures the number of data points in the training data is equivalent to the proportions in the original data set. (Strata below 10% of the total are pooled together.)

# Create a training data set with the training() function
# Pulls from training and testing sets created by initial_split()
multiVar_small_train <- training(multiVar_small_split)
multiVar_small_test <- testing(multiVar_small_split)
# Examine the environment to be sure # of observations looks like the 75/25 split. 3199:1066.

# Create a separate dataset of available IDs that were not used in the training dataset.
notTrain <- aoc %>% # all COMIDS from StreamCat data, sampled or not
  filter(!ID %in% aoc$ID) # Removing sites used to train the model. n = 140,097

count_optimized <- paste0('n = ',nrow(multiVar_small_roughfix))

skim(multiVar_small_roughfix)
```
#### Kitchen Sink Model
#### LINEAR DATA
```{r}
# Step Three - Kitchen Sink model -----------------------------------------
# Random forest -- 
# a decision tree model, using predictors to answer dichotomous questions to create nested splits.
# no pruning happens - rather, multiple trees are built (the forest) and then you are looking for consensus across trees
# training data goes down the tree and ends up in a terminal node.
# if testing data goes down the same route, then this upholds our conclusions. Or, if it goes awry, this allows us to look for patterns in how it goes awry.

set.seed(2) # assures the data pulled is random, but sets it for the run below (makes outcome stable)
myrf <- randomForest(y = multiVar_small_roughfix$effect_f, # dependent variable
  x = multiVar_small_roughfix %>%
    dplyr::select(-effect_f), # selecting all predictor variables
  importance = T, # how useful is a predictor in predicting values (nothing causal)
  proximity = T, 
  na.action = na.roughfix,
  ntrees = 100) # 500 trees default.

myrf # examine the results.
```
~20% error rate. Let's compare this with the same model with log-transformed values.

##### Predictor Selector/Recursive Feature Elimination
```{r}
# Using caret to select the best predictors
# What are the parameters you want to use to run recursive feature elimination (rfe)?
my_ctrl <- rfeControl(functions = rfFuncs,
                      method = "cv",
                      verbose = FALSE,
                      returnResamp = "all")

# rfe = recursive feature elimination
# THIS STEP TAKES FOR-EV-ER!!!
set.seed(22)
my_rfe <- rfe(y = multiVar_small_roughfix$effect_f, # set dependent variable
              x = multiVar_small_roughfix %>% 
                dplyr::select(-effect_f),
              rfeControl = my_ctrl,
               size = c(1:2, 4, 6, 8, 10, 12, 14, 15))#,
             # na.action = na.roughfix())

# sets how many variables are in the overall model
#               # I have 13 total possible variables, so I've chosen increments of 3 to look at.
#               rfeControl = my_ctrl,
#               na.action = na.roughfix,
#               testX = multiVar_small_test %>% dplyr::select(-effect_f),
#               testY = multiVar_small_test$effect_f)

# can you make your model even simpler?
# the following will pick a model with the smallest number of predictor variables based on the tolerance ("tol") that you specify (how much less than the best are you willing to tolerate?)

#inspect
my_rfe
```

```{r}
trellis.par.set(caretTheme())
#visualize
plot(my_rfe,type = c("g", "o"))
```

```{r}
my_size <- pickSizeTolerance(my_rfe$results, metric = "Accuracy", tol = 5, maximize = TRUE)
# higher tol (~10) gives you less variables
# lower tol (~1) gives you more variables - "I'd like the simplest model within 1% of the best model."
pickVars(my_rfe$variables, size = my_size)
```
The pickSizeTolerance determines the absolute best value then the percent difference of the other points to this value. This approach can produce good results for many of the tree based models, such as random forest, where there is a plateau of good performance for larger subset sizes. For trees, this is usually because unimportant variables are infrequently used in splits and do not significantly affect performance.

Just use best predictors.
```{r}
set.seed(2) # assures the data pulled is random, but sets it for the run below (makes outcome stable)
myrf_optimized <- randomForest(y = multiVar_small_train$effect_f, # dependent variable
  x = multiVar_small_train %>%
    dplyr::select(c(organism.group, exposure.duration.d, lvl1_f, dose.um3.mL.master)), # selecting all predictor variables
  importance = T, # how useful is a predictor in predicting values (nothing causal)
  proximity = T, 
  na.action = na.roughfix,
  ntrees = 100) # 500 trees default.

myrf_optimized # examine the results.
```
```{r}
varImpPlot(myrf_optimized)
```
```{r}

optimized_fitted <- predict(myrf_optimized, multiVar_small_test %>% dplyr::select(-effect_f), OOB = TRUE, type ="response")
misClasificError_optimized <- mean(optimized_fitted != multiVar_small_test$effect_f)
accuracy_optimized <- paste0('Accuracy: ', round(100*(1 - misClasificError_optimized), 2), '%')
accuracy_optimized
```

```{r}
df1 <- aoc_z %>% 
  dplyr::select(c(organism.group, exposure.duration.d, lvl1_f, dose.um3.mL.master, effect_10)) %>% 
  drop_na()

m1 <- glm(effect_10 ~ organism.group + exposure.duration.d + lvl1_f + dose.um3.mL.master, 
   data = df1, na.action = na.omit, family = "binomial")
summary(m1)
```

```{r}
require(ggeffects)
m1_g <- ggeffects::ggpredict(m1, terms = "dose.um3.mL.master")
plot(m1_g)
```
### Missing Value Imputation by RF

The algorithm starts by imputing NAs using na.roughfix. Then randomForest is called with the completed data. The proximity matrix from the randomForest is used to update the imputation of the NAs. For continuous predictors, the imputed value is the weighted average of the non-missing obervations, where the weights are the proximities. For categorical predictors, the imputed value is the category with the largest average proximity. This process is iterated iter times.

Note: Imputation has not (yet) been implemented for the unsupervised case. Also, Breiman (2003) notes that the OOB estimate of error from randomForest tend to be optimistic when run on the data matrix with imputed values.
```{r}
# impute values
#drop NA's in response
multiVar_small_noNa <- multiVar_small %>% drop_na(effect_f)

#impute
set.seed(111)
multiVar_small_rfImpute <- rfImpute(data = multiVar_small_noNa, 
                                    effect_f ~., #response value, cannot contain NA's
                                    iter = 4,
                                    ntree =75)

# Create calibration and validation splits with tidymodels initial_split() function.
set.seed(4)
multiVar_small_split_imputed <- multiVar_small_rfImpute %>%
  initial_split(prop = 0.75, strata = polymer) # splits data into training and testing set.
# default is 3/4ths split (but 75% training, 25% testing).
# Stratification (strata) = grouping training/testing sets by region, state, etc.
# Using the "strata" call ensures the number of data points in the training data is equivalent to the proportions in the original data set. (Strata below 10% of the total are pooled together.)

# Create a training data set with the training() function
# Pulls from training and testing sets created by initial_split()
multiVar_small_train_imputed <- training(multiVar_small_split_imputed)
multiVar_small_test_imputed <- testing(multiVar_small_split_imputed)
# Examine the environment to be sure # of observations looks like the 75/25 split. 3199:1066.

# Create a separate dataset of available IDs that were not used in the training dataset.
notTrain <- aoc %>% # all COMIDS from StreamCat data, sampled or not
  filter(!ID %in% aoc$ID) # Removing sites used to train the model. n = 140,097

count_optimized_imputed <- paste0('n = ',nrow(multiVar_small_rfImpute))
